In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-24 paper/main.tex

\newcommand{\rami}[1]{\textcolor{blue}{#1}}
\newcommand{\mer}[1]{\textcolor{magenta}{#1}}
\newcommand{\mar}[1]{\textcolor{violet}{#1}}
\newcommand{\mk}[1]{\textcolor{orange}{#1}}
\newcommand{\discuss}[1]{\textcolor{red}{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{#1}}
\newcommand{\draft}[1]{\textcolor{lightgray}{#1}}
%\newcommand{\brain}{\Omega_{\rm{PAR}}} 
%\newcommand{\sas}{\Omega_{\rm CSF}}
%\newcommand{\spinal}{\Gamma_{\rm SAS}}
%\newcommand{\skull}{\Gamma_{\rm{skull}}}
%\newcommand{\gin}{g_{\rm{influx}}}

@@

%\affil[+]{these authors contributed equally to this work}

%\keywords{Keyword1, Keyword2, Keyword3}

%% The problem we are addressing here is that we would like to be able to
%% understand, integrate and quantify the role of different mechanisms -
%% including effects associated with individual variations - on molecular
%% enrichment and clearance of human intracranial spaces. And we think of
%% this as a step on the way to predict tracer enrichment after
%% intrathecal injection in a personalized medicine context. Which is
%% important because of drug delivery/treatment in cancer, diagnostics
%% for CSF and dementia disorders, and clearance in neurodegenerative
%% disease.

@@

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
\thispagestyle{empty}

@@

\iffalse
\paragraph*{Abbreviations}
\begin{itemize}
\item
  Subarachnoid space (SAS)
\item
  Cerebrospinal fluid (CSF)
\item
  Perivascular space (PVS), perivascular spaces (PVSs)
\end{itemize}
\fi

@@

%% \begin{itemize}
%% \item
%%   Need better understanding of transport of solutes in intracranial spaces.
%% \item
%%   Need to integrate separate observations e.g. on diffusion in
%%   neuropil, PVS flow, net flow, effect of dispersion
%% \item
%%  Need to transfer and integrate insights from animal models,
%%   computational models and theoretical insights into the human case. 
%% \item
%%  Controversies need to be resolved.
%% \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
%% \mer{MER: Target ~3100 words in 6 subsections}

%% \mer{MER: Results #1 is about 500 words}
\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure1.png}
     \caption{A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}
(\Cref{fig:results1}D). Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

% \mer{MER: Results #2 is about 600 words}
\subsection*{Reduced CSF pulsatility strongly shifts intracranial enrichment patterns}
\begin{figure}%[h!]
\centering 
\includegraphics[width = 0.95\textwidth]{figures/figure2.png}
\caption{\textbf{The balance between CSF pulsatility and production shapes molecular enrichment patterns.}
A) CSF spaces: the third and fourth ventricle (V3\&4), lateral ventricles (LV) and SAS; 
B) Boundaries and interfaces: the spinal subarachnoid space (SSAS), upper and lower arachnoid membrane (AM-U and AM-L), pia (Pia) and lateral ventricles (LV); 
C) Schematic of the CSF flow model; 
D) CSF flow induced by CSF production with outflow allowed across the
upper convexity (AM-U); 
}
\label{fig:csf}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Histograms of the production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4) in terms of the relative frequency of the mean flow speed in each computational cell weighted by its volume; 
F) CSF flow induced at peak systolic blood inflow; 
G) Cardiac-induced dispersion factor $R_c$;
H) CSF flow induced at peak respiratory expansion; 
I) Respiration-induced dispersion factor $R_r$, $D = (1 + R_c + R_r) D^{\rm Gad}$; J--M): Histograms of the cardiac-induced flow velocities (J),
cardiac-induced dispersion factors (K),
respiratory flow velocities (L), respiratory dispersion factor (M);
N--Q): sagittal, coronal and axial planes of tracer concentration after 12h for different model variations: baseline (N), no CSF production (O), low dispersion (P), and high dispersion (Q).
}
\end{figure}

@@

% \mer{MER: Results #3 is about 800 words}
\subsection*{Perivascular flow shapes and accelerates molecular enrichment}
\label{sec:pvs_flow_results}

@@

We study the effect of such rapid PVS flow on intracranial molecular
enrichment by comparing the more conservative baseline model and a
high PVS flow model, where the former still includes the net flow
contributions from CSF production and the cardiac pulse wave, while
the latter additionally includes the ultraslow vasomotion
contribution. In both models, tracers were first observed at the basal
artery after 48 minutes (\Cref{fig:pvs}F), defined by their first-time
arrival (FTA), the time at which the concentration first exceeds $0.1$
mmol/l. At all upstream locations along the middle cerebral arteries
(MCAs) and anterior cerebral artery (ACA), we found substantially
reduced FTAs with higher PVS flow (\Cref{fig:pvs}G), up to 2.5 hours
earlier in the left M2 segment of the MCA. PVS concentrations peaked
before the surrounding tissue (0:48h vs 1:12h at the MCA and 1:00h vs
1:48h at MCA2) with high PVS flow, while these peaks occurred nearly
simultaneously in the baseline model. The earlier appearance and the
delay in time-to-peak between the PVS and surrounding tissue clearly
indicates the directionality of tracer enrichment -- it first arrives
in the PVS and subsequently spreads into the tissue and SAS
(\Cref{fig:pvs}H, I), and especially so with higher PVS flow. These
observations also hold on an aggregated level: computing the total
amount of tracer in the PVS as a function of the distance to the
arterial network roots, we find accelerated tracer transport with
higher PVS flow, especially at earlier time points (2--9 hours)
(\Cref{fig:pvs}J). Moreover, differences in tracer transport along the
PVS translate to altered enrichment patterns on the whole-organ
scale. For instance, after 4-6 hours, the faster-moving tracer in the
PVS is clearly visible in the adjacent spaces to the ACA, MCA, and
other arteries (\Cref{fig:pvs}L). \mer{One more sentence on effect of
  high PVS flow on tracer enrichment outside the PVS.} In contrast,
there are no clear signs of early enrichment surrounding the PVS in
the baseline model (\Cref{fig:pvs}K).

A natural question is whether early PVS enrichment could be the result
of increased dispersion rather than net flow in the
PVS\cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. Interestingly,
even increasing the dispersion factor by $100 \times$ in the PVS,
induced only minor changes in the global spreading rate compared to the
baseline (\Cref{fig:pvs}G, M), thus indicating a negative answer to
this question.
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}}
    \label{fig:pvs}
\end{figure}
\begin{figure}
\ContinuedFloat
  \caption{\textbf{(cont.)}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production (cf.~\Cref{fig:csf}D) -- histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) Estimated net PVS flow induced by pulse wave peristaltic pumping;
    E) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping;
    F) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    G) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);
    H, I) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));
    J) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    K--M) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentrations \discuss{below threshold masked}, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red).}
\end{figure}

% \mer{MER: Results #4 is about 500 words}
\subsection*{Structural versus functional compartmentalization of perivascular spaces}

@@

At the basal artery (BA), tracer appears within one hour in the surrounding CSF in both models (\Cref{fig:compartmentalization}C, E). With a more permeable PVS-CSF interface, tracer quickly crosses from the CSF into the PVS, resulting in similar peak concentrations of 25\,mmol/l in the PVS and at its outer surface, while values up to 50\,mmol/l are attained in the vicinity. In contrast, the low permeability model exhibits substantially lower perivascular tracer enrichment, with a maximum concentration of 10\,mmol/l after approximately two hours.
%The high permeability model allows for more tracer crossing into the PVS, leading to higher concentrations further downstream. Notably, crossover into the parenchyma is also observed, most clearly at 4 hours.
For the middle cerebral arteries (MCA-R and MCA-L), tracer appears
first in the PVS in both models (\Cref{fig:compartmentalization}D, E),
though with higher concentrations in the high permeability model (up
to 24 mmol/l). We also observe that the higher permeability allows the
tracer to leak out of the PVS and spread within the Sylvian fissure to
a greater extent. \discuss{With a small delay, more tracer appears via
  the CSF pathway (\Cref{fig:compartmentalization}F, peak time
  difference).} At the anterior cerebral artery (ACA-A2), the first tracer arrival in the PVS occurs after 3 hours, with a peak at 5 hours. The concentration peak in the PVS is tailed by a higher peak in the CSF (\Cref{fig:compartmentalization}E), indicating tracer arrival through a second pathway; i.e., directly through the CSF-filled space outside of the PVS. Similar patterns are observed for ACA-A3.

@@

% \mer{MER: Target Results 5 300 words}
\subsection*{Enlarged PVSs delay periarterial and intracranial molecular enrichment}

@@

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to \discuss{120} min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6h, while the mean \discuss{PVS}
concentration drops slightly (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
    \caption{\textbf{PVS enlargement reduces net PVS flow and delays tracer transport}
    A) Modelling dilated PVS by extending the outer PVS boundaries;
    B) mean and max PVS flow velocities with normal/control and dilated PVS;
    C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
    E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF)); 
    F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
    G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
    H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
    }
    \label{fig:enlargement}
\end{figure}

% \mer{MER: Target Results 6 400 words}
%\subsection*{The perivenous contribution to molecular clearance of CSF tracers}

%\begin{itemize}
%\item Look at clearance phase 12h-24h-48h-72h.
%\item Look at concentration in the veins
%\item Look at concentration in the parenchyma
%\item Consider low-pressure sink for the perivenous spaces.
%\end{itemize}

%% \subsection*{Clearance of intra-parenchymal solutes}

%% \begin{itemize}
%% \item
%%   Clearance of a molecular concentration in the parenchyma, different
%%   molecular sizes - ranging from Gadubutrol to amyloid-$\beta$/$\tau$.
%% \end{itemize}

@@

% Variations with disease states
%cohort variations, individual variations, physiological factor
%variations. Enlarged PVS reduce transport at the whole cerebral scale.

>>>>

\newcommand{\rami}[1]{\textcolor{blue}{#1}}
\newcommand{\mer}[1]{\textcolor{magenta}{#1}}
\newcommand{\mar}[1]{\textcolor{violet}{#1}}
\newcommand{\mk}[1]{\textcolor{orange}{#1}}
\newcommand{\discuss}[1]{\textcolor{red}{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{#1}}
\newcommand{\draft}[1]{\textcolor{lightgray}{#1}}

@@

@@

\flushbottom
\maketitle

\thispagestyle{empty}

@@

@@

\section*{Results}

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure1.png}
     \caption{A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}
(\Cref{fig:results1}D). Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

\subsection*{Reduced CSF pulsatility strongly shifts intracranial enrichment patterns}
\begin{figure}%[h!]
\centering 
\includegraphics[width = 0.95\textwidth]{figures/figure2.png}
\caption{\textbf{The balance between CSF pulsatility and production shapes molecular enrichment patterns.}
A) CSF spaces: the third and fourth ventricle (V3\&4), lateral ventricles (LV) and SAS; 
B) Boundaries and interfaces: the spinal subarachnoid space (SSAS), upper and lower arachnoid membrane (AM-U and AM-L), pia (Pia) and lateral ventricles (LV); 
C) Schematic of the CSF flow model; 
D) CSF flow induced by CSF production with outflow allowed across the
upper convexity (AM-U); 
}
\label{fig:csf}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Histograms of the production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4) in terms of the relative frequency of the mean flow speed in each computational cell weighted by its volume; 
F) CSF flow induced at peak systolic blood inflow; 
G) Cardiac-induced dispersion factor $R_c$;
H) CSF flow induced at peak respiratory expansion; 
I) Respiration-induced dispersion factor $R_r$, $D = (1 + R_c + R_r) D^{\rm Gad}$; J--M): Histograms of the cardiac-induced flow velocities (J),
cardiac-induced dispersion factors (K),
respiratory flow velocities (L), respiratory dispersion factor (M);
N--Q): sagittal, coronal and axial planes of tracer concentration after 12h for different model variations: baseline (N), no CSF production (O), low dispersion (P), and high dispersion (Q).
}
\end{figure}

@@

\subsection*{Perivascular flow shapes and accelerates molecular enrichment}
\label{sec:pvs_flow_results}

@@

We study the effect of such rapid PVS flow on intracranial molecular
enrichment by comparing the more conservative baseline model and a
high PVS flow model, where the former still includes the net flow
contributions from CSF production and the cardiac pulse wave, while
the latter additionally includes the ultraslow vasomotion
contribution. In both models, tracers were first observed at the basal
artery after 48 minutes (\Cref{fig:pvs}F), defined by their first-time
arrival (FTA), the time at which the concentration first exceeds $0.1$
mmol/l. At all upstream locations along the middle cerebral arteries
(MCAs) and anterior cerebral artery (ACA), we found substantially
reduced FTAs with higher PVS flow (\Cref{fig:pvs}G), up to 2.5 hours
earlier in the left M2 segment of the MCA. PVS concentrations peaked
before the surrounding tissue (0:48h vs 1:12h at the MCA and 1:00h vs
1:48h at MCA2) with high PVS flow, while these peaks occurred nearly
simultaneously in the baseline model. The earlier appearance and the
delay in time-to-peak between the PVS and surrounding tissue clearly
indicates the directionality of tracer enrichment -- it first arrives
in the PVS and subsequently spreads into the tissue and SAS
(\Cref{fig:pvs}H, I), and especially so with higher PVS flow. These
observations also hold on an aggregated level: computing the total
amount of tracer in the PVS as a function of the distance to the
arterial network roots, we find accelerated tracer transport with
higher PVS flow, especially at earlier time points (2--9 hours)
(\Cref{fig:pvs}J). Moreover, differences in tracer transport along the
PVS translate to altered enrichment patterns on the whole-organ
scale. For instance, after 4-6 hours, the faster-moving tracer in the
PVS is clearly visible in the space adjacent to the PVS of the ACA, MCA, and other arteries (\Cref{fig:pvs}L). In contrast,
there are no clear signs of early enrichment surrounding the PVS in
the baseline model (\Cref{fig:pvs}K).

A natural question is whether early PVS enrichment could be the result
of increased dispersion rather than net flow in the
PVS\cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. Interestingly,
even increasing the dispersion factor by $100 \times$ in the PVS,
induced only minor changes in the global spreading rate compared to the
baseline (\Cref{fig:pvs}G, M), thus indicating a negative answer to
this question.
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}}
    \label{fig:pvs}
\end{figure}
\begin{figure}
\ContinuedFloat
  \caption{\textbf{(cont.)}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production (cf.~\Cref{fig:csf}D) -- histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) Estimated net PVS flow induced by pulse wave peristaltic pumping;
    E) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping;
    F) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    G) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);
    H, I) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));
    J) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    K--M) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red).}
\end{figure}

\subsection*{Structural versus functional compartmentalization of perivascular spaces}

@@

At the basal artery (BA), tracer appears within one hour in the surrounding CSF in both models (\Cref{fig:compartmentalization}C, E). With a more permeable PVS-CSF interface, tracer quickly crosses from the CSF into the PVS, resulting in similar peak concentrations of 25\,mmol/l in the PVS and at its outer surface, while values up to 50\,mmol/l are attained in the vicinity. In contrast, the low permeability model exhibits substantially lower perivascular tracer enrichment, with a maximum concentration of 10\,mmol/l after approximately two hours.
%The high permeability model allows for more tracer crossing into the PVS, leading to higher concentrations further downstream. Notably, crossover into the parenchyma is also observed, most clearly at 4 hours.
For the middle cerebral arteries (MCA-R and MCA-L), tracer appears
first in the PVS in both models (\Cref{fig:compartmentalization}D, E),
though with higher concentrations in the high permeability model (up
to 24 mmol/l). We also observe that the higher permeability allows the
tracer to leak out of the PVS and spread within the Sylvian fissure to
a greater extent. With a small delay, more tracer appears via
  the CSF pathway (\Cref{fig:compartmentalization}D). At the anterior cerebral artery (ACA-A2), the first tracer arrival in the PVS occurs after 3 hours, with a peak at 5 hours. The concentration peak in the PVS is tailed by a higher peak in the CSF (\Cref{fig:compartmentalization}E), indicating tracer arrival through a second pathway; i.e., directly through the CSF-filled space outside of the PVS. Similar patterns are observed for ACA-A3.

@@

\subsection*{Enlarged PVSs delay periarterial and intracranial molecular enrichment}

@@

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to 96 min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6 and 12 h, while the mean PVS
concentration is initially lower, but later exceeds the baseline model (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
    \caption{\textbf{PVS enlargement reduces net PVS flow and delays tracer transport}
    A) Modelling dilated PVS by extending the outer PVS boundaries;
    B) mean and max PVS flow velocities with normal/control and dilated PVS;
    C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
    E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF)); 
    F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
    G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
    H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
    }
    \label{fig:enlargement}
\end{figure}

@@

% Not discussed variations with disease states

% Limitations
In terms of limitations, our primary constraint modelling-wise is the
geometric representation of the vascular (and therefore also
perivascular) networks. Concretely, we account for no vessels below
the original MRI resolution limits, and thus only a part of the
cerebral surface vasculature and effectively not the cortical
vessels. In addition, and again due to lack of resolution, there is
uncertainty associated with the location of the vessel segments
relative to the CSF spaces and parenchyma, with some vessels
surrounded by both domains and some vessels completely entering the
parenchyma before returning to the SAS. Higher resolution MRI data
(T1w and e.g. time-of-flight, QSM or other modalities) in more
subjects would immediately allow for more detailed studies of this
aspect. As a result of the lack of an accurate representation of the
cortical and subcortical white matter vasculature, we have considered
a conservative transport model within the parenchyma, modelling
extracellular diffusion alone. To compensate, we have focused on
reporting detailed predictions in the CSF spaces and surface PVS after
not more than 24 hours. In particular, we note that, we observe tracer
accumulation at the perivascular network ends after \fixme{24} hours,
similar to that previously observed for
microspheres~\cite{bedussi2018paravascular, mestre2018flow}, which
here can be interpreted as an artifact of the incomplete network
creating up a barrier for tracer movement. Previously, we have found
that enrichment within the parenchyma is well-represented by $\approx
3\times$ enhanced diffusion augmented by local clearance e.g.~across
the blood-brain barrier or diffusion-convection with a complex pattern
of spatially-varying convective velocities of 1-10 \textmu
m/min~\cite{vinje2023human}. We will incorporate such transport
mechanisms within the parenchyma in future work. Finally, we also note
that we simulate the intracranial distribution of a tracer
concentration appearing at the craniocervical junction rather than as
an intrathecal injection. We therefore overestimate the amount of
tracer entering into the intracranial spaces; however, due to the
linearity of the mathematical model, it is valid to directly interpret
the in-silico concentrations relative to the total amount of tracer.

% Outlook
In conclusion, our findings transfer insights from experimental
studies and theoretical analysis to the in-silico human setting,
reconcile seemingly conflicting observations in particular relating to
directionality of perivascular flow, and integrate different physical
mechanisms across across spatial and temporal scales. The complete
simulation pipeline is openly available, including interactive
visualization of simulation results for all model
variations~\cite{ZENODO}. Looking ahead, this platform establishes a
foundation for in-silico studies of molecular movement within the
human brain environment, such as tailored predictions of intrathecal
chemotherapy delivery or personalized diagnostics of brain clearance
capacity.
